```{r setup, echo=FALSE}
library(knitr)
opts_chunk$set(dev='CairoPNG')

saveCytoscapeSVG <- function(before, options, envir){
  heightStr <- ""
  widthStr <- ""
  if (!is.null(options$out.height)){
    heightStr <- paste('height="', options$out.height, '"', sep="")
  }
  if (!is.null(options$out.width)){
    widthStr <- paste('width="', options$out.width, '"', sep="")
  }
  if (!before){
    if (exists("cw")){
      svgFile <- paste(options$label, ".svg", sep="")
      imgURL <- paste("figure", svgFile, sep="/")
      savePath <- file.path(getwd(), "figure", svgFile)
      saveImage(cw, savePath, "svg")
      paste('<img src="', imgURL, '" title="', options$label, '" alt="', options$label, '" ', heightStr, ' ', widthStr, '/>', sep="")
    }
  }
}

knit_hooks$set(saveCytoscape=saveCytoscapeSVG)
```


# A Review of R, Cytoscape and Python Integration for Data Visualization and Exploration

[`R`](http://cran.r-project.org/) is a popular data analysis and visualization environment, with a large number of user contributed packages, and the ability for interested individuals to easily contribute their own packages. 

[`Cytoscape`](cytoscape.org) is a popular network visualization software.

[`Python`](python.org) is an alternative popular data analysis environment.

All of the code and data used is available [on github](https://github.com/rmflight/wg_vis_presentation).

## Why Script Data Visualization and Exploration??

There are many tools people can use to interact with their data, with Microsoft Excel (and other spreadsheet packages) being very popular. However, the number of different ways that data can be easily plotted or manipulated is very limiting in spreadsheet packages. Other commercial visualization packages such as Tableau and Spotfire are more powerful than spreadsheet software, and enable greater control, but **reproducing** a plot later can be difficult if one does not record the exact set of options and tick boxes selected in producing a visualization.

I **believe** that scripting languages such as `R` and `Python` that allow rather easy data manipulation and visualization provide the best of both worlds. For the `Python` section of this, please go [here](http://nbviewer.ipython.org/github/rmflight/wg_vis_presentation/blob/master/matplotlib%20demo.ipynb)

## Why R, ggplot2

`R` is useful because of the various statistical utilities available. `R` has a set of base graphics utilities that allow one to do various types of plots. As an example, we can create a scatterplot of two variables:

```{r scatter2}
var1 <- rnorm(100)
var2 <- rnorm(100, 1)
plot(var1, var2)
```

Or do a histogram of one of the variables:

```{r histogram}
hist(var1)
hist(var2)
```

But doing much else in base graphics can get rather tedious and difficult, especially when you want multiple plots.

As a simple example, lets imagine that we have two replicates of each variable that we want to examine.

```{r tworeps}
varData <- data.frame(var1=c(rnorm(100), rnorm(100)), var2=c(rnorm(100, 1), rnorm(100,1)), sample=rep(c(1,2), each=100))
```

Using base graphics, we can easily plot the two variables against each other:

```{r plotRepsBase}
plot(varData$var1, varData$var2)
```

But if we want to color each sample differently on the same graphic (a common enough thing), things quickly get more complicated:

```{r plotRepsBasebySample}
plot(varData$var1[varData$sample==1], varData$var2[varData$sample==1], col="blue")
points(varData$var1[varData$sample==2], varData$var2[varData$sample==2], col="red")
```

We are no longer using the base `plot` command, but a sub-part of it. And plumbing it's depths quickly becomes tedious.

## GGPlot2

Enter [`ggplot2`](http://ggplot2.org/). It is an `R` implementation of the [*grammar of graphics*](http://www.amazon.com/Grammar-Graphics-Statistics-Computing/dp/0387245448). To better understand it's design and how it treats entries and variables, I recommend reading Hadley Wickham's [Tidy Data](http://vita.had.co.nz/papers/tidy-data.pdf) manuscript.

Using `ggplot2`, our previous graph becomes rather trivial.

```{r plotRepsSampleGGplot}
library(ggplot2)
varData$sample <- factor(varData$sample)
ggplot(varData, aes(x=var1, y=var2, colour=sample)) + geom_point()
```

We can also easily create multiple sub-plots that are broken out the by **sample** variable:

```{r plotsampleOut}
ggplot(varData, aes(x=var1, y=var2)) + geom_point() + facet_grid(. ~ sample)
```


In it's current form, we can easily plot summary histograms of each variable:

```{r plotHistggplot}
ggplot(varData, aes(x=var1, fill=sample)) + geom_bar(position="identity", alpha=0.5)
ggplot(varData, aes(x=var1, fill=sample)) + geom_density(alpha=0.5)
ggplot(varData, aes(x=var1)) + geom_bar() + facet_grid(sample ~ .)
```

If we want to do a bigger overview of summary statistics, we can just reshape our data a little bit:asdfasdf

```{r plotSummary}
library(reshape2)
varData2 <- melt(varData, measure.vars=c("var1", "var2"))

ggplot(varData2, aes(x=value, fill=sample)) + geom_bar(position="identity", alpha=0.5) + facet_grid(. ~ variable)
ggplot(varData2, aes(x=value)) + geom_bar() + facet_grid(sample ~ variable)
```

## Layers & Customization

The other major strength of `ggplot` is the ability to layer different graphical on top of each other, as well as customizing plots.

### Layers

Layers are the individual plot elements that are added to a given plot. For example, we can do:

```{r layers}
p <- ggplot(varData, aes(x=var1, y=var2))
p + geom_point()
p + stat_smooth()
p + geom_point() + stat_smooth()
```

### Customization

We can customize just about any aspect of a plot that we want.

From the color of the points plotted:

```{r customizePoints}
p + geom_point(color="red")
```

To the axis labels:

```{r axisLabels}
p + geom_point() + labs(x="Important Data 1", y="Dependent Variable 1")
```

To the fonts, background, etc, or the elements of a particular `ggplot` **theme**:

```{r theme}
pPoint <- p + geom_point()
pPoint + theme(axis.text = element_text(colour = "red"))
pPoint + theme(axis.text = element_text(size = 14))
pPoint + theme(axis.title = element_text(color = "blue", size = 20))
```

There are a couple of built-in themes such as `theme_bw` and `theme_classic` alongside the default, and there is also a collection of alternative themes in the [`ggthemes`](https://github.com/jrnold/ggthemes) package.

## With Metabolomics Data

Metabolomics and network data are courtesy of Dmitry Grapov's [Creative Data Solutions](http://imdevsoftware.wordpress.com/2014/02/17/tutorials-statistical-and-multivariate-analysis-for-metabolomics/) metabolomic tutorial.

```{r loadData}
metabData <- read.table("pumpkin_tomatillo_data.csv",
                        sep=",", header=T, stringsAsFactors=FALSE)
colnames(metabData)[4:151] <- seq(1, 148)

metab2 <- melt(metabData, id.vars=seq(1,3), measure.vars=seq(4, 151))
varNameData <- read.table("metabID.csv", sep=",", header=T, stringsAsFactors=FALSE)

metab2$metabName <- rep(varNameData$name, each=24)
metabData <- metab2
newTreat <- rep("fresh", nrow(metabData))
newTreat[grep("lyophilized", metabData$Treatment)] <- "lyoph"
metabData$treat <- newTreat
head(metabData)
```

```{r outputData, eval=FALSE}
write.table(metabData, file="metabolomics_reshapedData.csv", sep=",", row.names=FALSE, col.names=TRUE)
```


Let's summarize the values for each type of sample. We will stop at a value of **5000** because 

```{r viewHistograms}
ggplot(metabData, aes(x=value, fill=treat)) + geom_bar(alpha=0.8, position="identity") + facet_grid(Species ~ .) + xlim(0, 5000)

ggplot(metabData, aes(x=value, fill=Species)) + geom_bar(alpha=0.8, position="identity") + facet_grid(treat ~ .) + xlim(0, 5000)
```

As an alternative to histograms, `ggplot` provides smoothed density estimates that can nicely capture the **shape** of a distribution.

```{r viewDensity}
ggplot(metabData, aes(x=value, fill=treat)) + geom_density(alpha=0.8) + facet_grid(Species ~ ., scales="free_y") + xlim(0, 5000)
ggplot(metabData, aes(x=value, fill=Species)) + geom_density(alpha=0.8) + facet_grid(treat ~ ., scales="free_y") + xlim(0, 5000)
```

We can also do the boxplot for all the metabolites, with a separate box for each species.

```{r plotAll}
ggplot(metabData, aes(x=metabName, y=value, fill=Species)) + geom_boxplot()
```

This is not that easily to decipher, but it was quite easy to do.


Based on the analysis presented in the tutorial, we can pick out some of the top changers:

```{r topChangers}
topChangers <- c(34, 62, 8, 17, 124, 100, 115, 9, 93, 123, 75, 71, 85, 33)
metabTop <- metabData[(metabData$variable %in% topChangers),]
ggplot(metabTop, aes(x=metabName, y=value, fill=Species)) + geom_boxplot()
ggplot(metabTop, aes(x=metabName, y=value, fill=Species)) + geom_boxplot() + scale_y_log10()
```

Finally, what is also quite nice about using `ggplot2` is the ability to add incrementally to a graph. In our previous example, we could have achieved the same result piecemeal like this:

```{r topChangersPiecemeal}
g <- ggplot(metabTop, aes(x=metabName, y=value, fill=Species)) + geom_boxplot()
print(g)
g + scale_y_log10()
```


# Networks

Networks are very useful things in many fields, due to the fact that we can easily come up with ways to define relationships between objects. These objects can be anything, but in the context of metabolomics they are often `genes`, `proteins`, and `metabolites`. The relationships between metabolites can be defined in a couple of different ways. This example we will define edges between metabolites by links in the `KEGG` database, and by their chemical similarity (tanimoto coefficient). All data was generated by `metamapR`.

The network graph will be constructed in `R`, and visualized in `Cytoscape`. This gives us the flexibility to manipulate the graph easily in `R` and save groups of relevant nodes, with the visual richness and interactive flexibility of `Cytoscape`.

This connection of `R` with `Cytoscape` uses the [`RCytoscape`](http://rcytoscape.systemsbiology.net/versions/current/index.html) `Bioconductor` package, and requires `Cytoscape` v2.8, and the `CytoscapeRPC` plugin. There is currently no way to do this with `Cytoscape` v3.0, however given that a `REST` api is being introduced to `Cytoscape`, it should be possible in the future (if we can persuade Paul Shannon to write a new `R` package).

```{r loadNetworkData}
library(graph)
nodeEdges <- read.table("biochemEdgeList.csv", header=T, sep=",", stringsAsFactors=FALSE)

# need character for node IDs
nodeEdges[,1] <- as.character(nodeEdges[,1])
nodeEdges[,2] <- as.character(nodeEdges[,2])
```

```{r createGraph}
allNodes <- unique(c(nodeEdges$source, nodeEdges$target))

metabNetwork <- graphNEL(nodes=allNodes, edgemode="directed")
metabNetwork <- addEdge(nodeEdges$source, nodeEdges$target, metabNetwork, nodeEdges$weight)
```

We also need the node and edge attributes, and to decide how they will be represented in the graph so that the types can be properly passed to Cytoscape. 

```{r nodeProperties}
nodeData <- read.table("node_attributes.csv", header=T, sep=",", stringsAsFactors=FALSE)
nodeData$Index <- as.character(nodeData$Index)

nodeDataDefaults(metabNetwork, "name") <- ""
attr(nodeDataDefaults(metabNetwork, "name"), "class") <- "STRING"
nodeDataDefaults(metabNetwork, "direction") <- ""
attr(nodeDataDefaults(metabNetwork, "direction"), "class") <- "STRING"
nodeDataDefaults(metabNetwork, "sig.dir") <- ""
attr(nodeDataDefaults(metabNetwork, "sig.dir"), "class") <- "STRING"
nodeDataDefaults(metabNetwork, "size.FC") <- 0
attr(nodeDataDefaults(metabNetwork, "size.FC"), "class") <- "DOUBLE"
nodeDataDefaults(metabNetwork, "size.log.FC") <- 0
attr(nodeDataDefaults(metabNetwork, "size.log.FC"), "class") <- "DOUBLE"
nodeDataDefaults(metabNetwork, "size.LV") <- 0
attr(nodeDataDefaults(metabNetwork, "size.LV"), "class") <- "DOUBLE"
nodeDataDefaults(metabNetwork, "log.mean.Pumpkin") <- 0
attr(nodeDataDefaults(metabNetwork, "log.mean.Pumpkin"), "class") <- "DOUBLE"
nodeDataDefaults(metabNetwork, "log.mean.Tomatillo") <- 0
attr(nodeDataDefaults(metabNetwork, "log.mean.Tomatillo"), "class") <- "DOUBLE"

edgeDataDefaults(metabNetwork, "weight") <- 0
attr(edgeDataDefaults(metabNetwork, "weight"), "class") <- "DOUBLE"
edgeDataDefaults(metabNetwork, "source") <- ""
attr(edgeDataDefaults(metabNetwork, "source"), "class") <- "STRING"

edgeData(metabNetwork, nodeEdges[,1], nodeEdges[,2], "source") <- nodeEdges$type


useNodes <- nodeData$Index %in% nodes(metabNetwork)
for (useAttr in c("name", "direction", "sig.dir", "size.FC", "size.log.FC", "size.LV", "log.mean.Pumpkin", "log.mean.Tomatillo")){
  
  nodeData(metabNetwork, nodeData$Index[useNodes], useAttr) <- nodeData[useNodes, useAttr]
}
```

Now that we've added all the nitty gritty details to our network, we can pass it to Cytoscape and start doing some visualization.

```{r loadCytoscape, saveCytoscape=TRUE, out.width="800px"}
library(RCytoscape)
cw <- new.CytoscapeWindow("metabolite similarity", graph=metabNetwork)
displayGraph(cw)
```

Not much to look at yet, we have to tell `Cytoscape` to layout the network.

```{r layoutNetwork, saveCytoscape=TRUE, out.width="800px"}
setLayoutProperties(cw, 'force-directed', list(edge_attribute='weight'))
layoutNetwork(cw, 'force-directed')
```

As it stands, we can't get a whole lot of information yet from this network. But lets start adding some visual properties to it and see what happens.

## Edges

Lets change the edge visuals based on the source. Maybe black for KEGG (biochemical) and gray for Tanimoto (similarity)?

```{r changeEdgeColors, saveCytoscape=TRUE, out.width="800px"}
edgeColors <- c('#000000', '#999999')
setEdgeColorRule(cw, 'source', c('KEGG', 'Tanimoto'), edgeColors, mode='lookup')
redraw(cw)
```

So now we can see that most of the edges are from the similarity measure.

## Nodes

### Colors

We have metabolites that have changes either in the up or down direction. It would be nice to at a glance see which is which.

```{r changeNodeColors, saveCytoscape=TRUE, out.width="800px"}
nodeColors <- c('#99FF00', '#FF9900')
setNodeColorRule(cw, 'direction', c('increase', 'decrease'), nodeColors, mode='lookup')
redraw(cw)
```

### Display Labels

Currently, we have to select a node to see it's metabolite name. We can use something else as the display label.

```{r changeLabel, saveCytoscape=TRUE, out.width="800px"}
setNodeLabelRule(cw, 'name')
redraw(cw)
```


## Select Nodes and Neighbors

We can select nodes based on particular properties in our original data. Such as those nodes that had large fold-changes, and their immediate neighbors:

```{r selectHiFC, saveCytoscape=TRUE, out.width="800px"}
hiFCNodes <- nodeData$Index[abs(nodeData$size.log.FC) >= 1.2]
selectNodes(cw, hiFCNodes)
```

```{r selectHiFCNeighbors, saveCytoscape=TRUE, out.width="800px"}
selectFirstNeighborsOfSelectedNodes(cw)
hiFCNodesNeighbors <- getSelectedNodes(cw)
hiFCNodesNeighborsData <- nodeData[nodeData$Index %in% hiFCNodesNeighbors,]
```

## Graph Properties

We can also use other `R` packages like `igraph` to analyze our network, and select particular groups of nodes based on the computed properties. In this example we will analyze the network for communities of nodes:

```{r graphNeighborhood}
library(igraph)

tmpNetwork <- metabNetwork
edgeDataDefaults(tmpNetwork, "source") <- NULL
edgeDataDefaults(tmpNetwork, "weight") <- NULL
nodeData(tmpNetwork, nodes(tmpNetwork), "name") <- nodes(tmpNetwork)
igNetwork <- igraph.from.graphNEL(tmpNetwork, name=TRUE)
igNetwork <- as.undirected(igNetwork)
eb <- edge.betweenness.community(igNetwork)

eb
```

What does this get us? A set of communities, with each node assigned to each community. We can examine them in `Cytoscape` by doing:

```{r examineCommunities}
allComms <- split(eb$names, eb$membership)
allComms <- allComms[order(sapply(allComms, length), decreasing=T)]
```

```{r community1, saveCytoscape=TRUE, out.width="800px"}
selectNodes(cw, allComms[[1]], preserve.current.selection=FALSE)
```

```{r community2, saveCytoscape=TRUE, out.width="800px"}
selectNodes(cw, allComms[[2]], preserve.current.selection=FALSE)
```

```{r community3, saveCytoscape=TRUE, out.width="800px"}
selectNodes(cw, allComms[[3]], preserve.current.selection=FALSE)
```

```{r community4, saveCytoscape=TRUE, out.width="800px"}
selectNodes(cw, allComms[[4]], preserve.current.selection=FALSE)
```

```{r}
deleteWindow(cw)
```

## Saving Plots

We often times want to save our visualization for later use in publications and reports. One important consideration is what format. PDF is frequently useful for hi-resolution lineart, but so is SVG (scalable vector graphics). If you have lots of points, SVG and PDF are not a good idea due to encoding each individual point. For those types of plots, you want to use bitmaps, such as PNG (preferred), TIFF (next up), or (shudder) JPG.

### General

To save just about any plot on any computer system that runs `R`, one can save graphics using the `Cairo` package relatively painlessly:

```{r}
library(Cairo)
```

```{r, eval=FALSE}
CairoPNG(file="testPNG.png")
plot(var1, var2)
dev.off()
```

### Networks

To save our network visualizations generated by `Cytoscape`, you should use the `saveImage` function. Again, unless there are a **lot** of nodes, you should probably use SVG or PDF, otherwise use PNG. Note that `saveImage` needs a full path on Linux.

```{r saveCytoscape, eval=FALSE}
saveImage(cw, file.name=file.path(getwd(), "testNetwork.svg"), image.type="svg")
```

## Resources and Links

### R

* [cran](http://cran.r-project.org)
* [R for Cats](http://rforcats.net)
* [Hadley Wickhams Advanced R Programming](http://adv-r.had.co.nz/)

### ggplot

* [Official ggplot2 documentation](http://docs.ggplot2.org/current/)
* [Noam Ross's ggplot2 tutorial](http://www.noamross.net/blog/2012/10/5/ggplot-introduction.html)
* [Stackoverflow questions on ggplot2](http://stackoverflow.com/questions/tagged/ggplot2)

### Cytoscape 

* [Cytoscape home page](http://cytoscape.org)
* [RCytoscape Project](http://bioconductor.org/packages/release/bioc/html/RCytoscape.html)


## System Information

```{r}
sessionInfo()
```

```{r}
purl("wg_visualization_ggplot_cytoscape_matplotlib.Rmd", quiet=TRUE)
```

