# Visualization WG: ggplot, cytoscape, matplotlib

Metabolomics data is from [Creative Data Solutions](http://imdevsoftware.wordpress.com/2014/02/17/tutorials-statistical-and-multivariate-analysis-for-metabolomics/) metabolomic tutorial.

## Why R, ggplot2

`R` is useful because of the various statistical utilities available. `R` has a set of base graphics utilities that allow one to do various types of plots. As an example, we can create a scatterplot of two variables:

```{r scatter2}
var1 <- rnorm(100)
var2 <- rnorm(100, 1)
plot(var1, var2)
```

Or do a histogram of one of the variables:

```{r histogram}
hist(var1)
hist(var2)
```

But doing much else in base graphics can get rather tedious and difficult, especially when you want multiple plots.

As a simple example, lets imagine that we have two replicates of each variable that we want to examine.

```{r tworeps}
varData <- data.frame(var1=c(rnorm(100), rnorm(100)), var2=c(rnorm(100, 1), rnorm(100,1)), sample=rep(c(1,2), each=100))
```

Using base graphics, we can easily plot the two variables against each other:

```{r plotRepsBase}
plot(varData$var1, varData$var2)
```

But if we want to do it by sample, things quickly get more complicated:

```{r plotRepsBasebySample}
plot(varData$var1[varData$sample==1], varData$var2[varData$sample==1], col="blue")
points(varData$var1[varData$sample==2], varData$var2[varData$sample==2], col="red")
```

We are no longer using the base `plot` command, but a sub-part of it. And plumbing it's depths quickly becomes tedious.

## GGPlot2

Enter `ggplot2`. It is an `R` implementation of the [*grammar of graphics*](http://www.amazon.com/Grammar-Graphics-Statistics-Computing/dp/0387245448). To better understand it's design and how it treats entries and variables, I would read Hadley Wickham's [Tidy Data](http://vita.had.co.nz/papers/tidy-data.pdf).

In `ggplot2`, our previous graph becomes rather trivial.

```{r plotRepsSampleGGplot}
library(ggplot2)
varData$sample <- factor(varData$sample)
ggplot(varData, aes(x=var1, y=var2, colour=sample)) + geom_point()
ggplot(varData, aes(x=var1, y=var2)) + geom_point() + facet_grid(. ~ sample)
```

In it's current form, we can easily plot summary histograms of each variable:

```{r plotHistggplot}
ggplot(varData, aes(x=var1, fill=sample)) + geom_bar(position="identity", alpha=0.5)
ggplot(varData, aes(x=var1, fill=sample)) + geom_density(alpha=0.5)
```

If we want to do a bigger overview of summary statistics, we can just reshape our data a little bit:

```{r plotSummary}
library(reshape2)
varData2 <- melt(varData, measure.vars=c("var1", "var2"))

ggplot(varData2, aes(x=value, fill=sample)) + geom_bar(position="identity", alpha=0.5) + facet_grid(. ~ variable)
ggplot(varData2, aes(x=value)) + geom_bar() + facet_grid(sample ~ variable)
```

## With Metabolomics Data

```{r loadData}
metabData <- read.table("pumpkin_tomatillo_data.csv",
                        sep=",", header=T, stringsAsFactors=FALSE)
colnames(metabData)[4:151] <- seq(1, 148)

metab2 <- melt(metabData, id.vars=seq(1,3), measure.vars=seq(4, 151))
varNameData <- read.table("metabID.csv", sep=",", header=T, stringsAsFactors=FALSE)

metab2$metabName <- rep(varNameData$name, each=24)
metabData <- metab2
```

